Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela

### 2024-12-02 agujournaltemplate.tex
monday afternoon corrections and results writing

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time-Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (about 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at 8 standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela), and from mid-1997 to the present, by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. The data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ data showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged. For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.

    For phytoplankton species identification by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all depth discrete measurements, the data was interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated and then grouped by Genus via the sum of interpolated counts.

    \subsection{Data analysis}    
    To give a general overview of the data and trends over the time series, we calculated yearly means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). The cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the dataset more homogeneous and more robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}. Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

    %RUE/BDEF ANALYSIS
    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling. $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
    %here I want to extend to glmm that correct for depth and seasonality for the final results

    \subsubsection{Clustering and NMDS}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of yearly aggregates of genera observed integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We used the Bray distance matrix in a 2-dimensional ordination, that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a 2-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify ranges of the predictors where significant changes occur in indicator variables, which signal a threshold for community turnover in the phytoplankton community. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. These analyses were conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, -u 10m wind component, the 21°C Isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which Gradient Forest methodology takes into account by implementing a conditional permutation of correlated predictors \cite{ellis_gradient_2012}. To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run. % see figure in supplementary with initial runs TODO

\section{Draft Results}
Our study focused on the phytoplankton microscopy data from the CARIACO Time Series, which consists of mostly monthly samplings taken at multiple depths between November 1995 and January 2017. We compared this data with nutrient, temperature and salinity data from Niskin bottle samples, as well as climate data and indices, focusing on the bottom-up drivers of variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available. 
% Figure 1
Normalized yearly means for all used environmental variables, as well as chlorophyll a and functional group counts, are shown in Figure \ref{fig:zscore}. The patterns emerging from -u 10m wind component, Temperature and the directly related 21°C Isotherm show clearly the years of positive anomalies of wind (red),  upwelling (blue, i.e. shallower depth of isotherm) and negative for temperature (blue) within the top 100 m. These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing oscillating patterns. Nitrate, and to some extent phosphate, seem to follow the upwelling anomalies, but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration towards the later part of the time series (between 2008 and 2013), which could be driven by surface runoff and reduced depletion via diminished Diatom abundances.

% Figure 2

@@

\begin{figure}
\label{fig:zscore}
\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. All data is available at monthly resolution (see next draft figure for raw data). Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\end{figure}

\begin{figure}
\label{fig:divts}
\begin{center}
\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries.pdf}
\end{center}
\caption{Overview over Chlorophyll and Diversity Variables over Depth}
\end{figure}

>>>>

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m. This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.

\subsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}. Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise. These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).

    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr. K. Fanning (University of South Florida, St. Petersburg, FL). Samples from eight cruises were analyzed by both laboratories. Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies. \cite{taylor_ecosystem_2012}. For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged. For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.

    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate. Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period. Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
    %from Mutshinda et al. 2013
    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}). After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.

@@

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}. All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}. %Note: Repo currently still set to private
    For all depth discrete measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.

    \subsection{Data analysis}    
    To give a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$). Cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}. Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

    %RUE/BDEF ANALYSIS
    To investigate the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity, and resource use efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index, and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling. $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
    %here I want to extend to glmm that correct for depth and seasonality for the final results

    \subsubsection{Clustering and NMDS}
    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth. Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}. We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps. To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations. The arrows indicate the gradient of a 2-dimensional linear trend surface. 
    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}. %Here I want to add "significant diferences calculation" using e.g. ANOVA.
    % detailed explanation of NMDS in Winder and Hunter 2008

    \subsubsection{Gradient Forest}
    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for community turnover in the phytoplankton community. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data. The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$). The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run. % see figure in supplementary with initial runs TODO

\section{Draft Results}
Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017. We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity. In total, we used a complete set of 254 monthly data points, on which all variables were available. 
% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing interannual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a marked reduction in biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.

@@

\begin{figure}
\label{fig:zscore}
\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
\caption{Heatmap of yearly average values, normalized for comparison as z-scores. All data is available at monthly resolution (see next draft figure for raw data). Red indicates a larger value, blue indicates a smaller value for that particular year. The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data. Diversity Indices are calculated based on microscopy counts at the Genus level.}
\end{figure}

\begin{figure}
\label{fig:divts}
\begin{center}
\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries_updated.pdf}
\end{center}
\caption{Overview over Chlorophyll and Diversity Variables over Depth}
\end{figure}
